From Reversible Programs to \\ Univalent Universes and Back

### 2021-10-30 Pi/paper/applications.tex
Say symmetric group instead of permutation group

In our development, we described a number of different representations of permutations: $\PiLang$ programs, lists of
transpositions, Lehmer codes, and automorphisms of finite sets. The equivalences between these representations allows us
to transfer theorems about permutations, proving them on the representations most suited for the particular task and
getting others for free.

@@

\noindent
Now, the question we want to answer is: is it possible to compose some number of (extended) 3-bit toffoli gates in some way, so as to replicate the
behaviour of a 4-bit toffoli gate?

Solving this problem using the circuit representation is difficult. However, we can observe that 3-bit toffoli gate
implements an even permutation, that is, a permutation consisting of an even number of transpositions. On the other
hand, 4-bit toffoli gate is an odd permutation. It is an easily proven lemma that composition of even permutations
always produces an even permutation, which answers the question to the negative.

We define a function $\parity : (c : X \iso Y) \to \Bool$ to compute a parity of a permutation defined by a combinator
$c$. It is implemented as the transformation from a $\PiLang$ program into a list of transposition, using $\evalh_{1}$
defined in~\cref{sec:equivalence}, and computing the parity on this representation, which means simply checking the
parity of the length of the list.

>>>>

In our development, we described a number of different representations of permutations: $\PiLang$, $\PiPlusLang$ and $\PiHatLang$ programs, lists of
transpositions, Lehmer codes, and automorphisms of finite sets. The equivalences between these representations allows us
to transfer theorems about permutations, proving them on the representations most suited for the particular task and
getting others for free.

@@

\noindent
Now, the question we want to answer is: is it possible to implement \Afun{toffoli₄} gate only using \Afun{toffoli₃} gates?

Solving this problem using the circuit representation is difficult. However, we can observe that all \Afun{toffoli₃} gates
implements even permutations, that is, permutations consisting of an even number of transpositions. On the other
hand, \Afun{toffoli₄} implements an odd permutation. It is an easily proven lemma that any composition of even permutations
always produces an even permutation, which answers the question to the negative.

Thus, define a function $\parity : (c : X \iso Y) \to \Bool$ to compute a parity of a permutation defined by a combinator
$c$. It is implemented first tranforming a $\PiLang$ program into a list of transpositions (using $\evalh_{1}$
defined in~\cref{sec:equivalence}), and then computing the parity on this representation, which amounts to a simple check of the length of the list.

### 2021-10-30 Pi/paper/discussion.tex
Say symmetric group instead of permutation group

Coherence and normalisation problems for monoids in constructive type theory using coherence for monoidal categories
were studied by \citet{beylinExtractingProofCoherence1996}. In HoTT, coherence for the free monoidal groupoid over a
groupoid and the proof of its universal property has been considered
by~\citet{piceghelloCoherenceMonoidalGroupoids2020}. Free commutative monoids in type theory have been studied by
\citet{gylterudMultisetsTypeTheory2020}, and in HoTT by
\citet*{choudhuryFinitemultisetConstructionHoTT2019,choudhuryFreeCommutativeMonoids2021}. The free symmetric monoidal
groupoid $\FSM[A]$ over a groupoid $A$ can be given by $\dsum*{X:\UFin}{A^{X}}$, or it can be presented as an algebraic
2-theory using 1-HITs. These HITs and the proof of their universal property have been considered by
\citet*{piceghelloCoherenceSymmetricMonoidal2019,choudhuryFinitemultisetConstructionHoTT2019}. The proof of the
universal property of $\FSM$ is asserted by appealing to Mac Lane's coherence theorem for symmetric monoidal categories,
and using the fact that the finite symmetric group $\Sn$ encodes the permutation group on a finite set. The existence of
the proof is folklore. We have produced a proof and formalised it in constructive type theory. A stronger universal
property that $\BFin$ is biinitial in $\term{RigCat}$, is in~\cite{elguetaGroupoidFiniteSets2021}.

>>>>

Coherence and normalisation problems for monoids in constructive type theory using coherence for monoidal categories
were studied by \citet{beylinExtractingProofCoherence1996}. In HoTT, coherence for the free monoidal groupoid over a
groupoid and the proof of its universal property has been considered
by~\citet{piceghelloCoherenceMonoidalGroupoids2020}. Free commutative monoids in type theory have been studied by
\citet{gylterudMultisetsTypeTheory2020}, and in HoTT by
\citet*{choudhuryFinitemultisetConstructionHoTT2019,choudhuryFreeCommutativeMonoids2021}. The free symmetric monoidal
groupoid $\FSM[A]$ over a groupoid $A$ can be given by $\dsum*{X:\UFin}{A^{X}}$, or it can be presented as an algebraic
2-theory using 1-HITs. These HITs and the proof of their universal property have been considered by
\citet*{piceghelloCoherenceSymmetricMonoidal2019,choudhuryFinitemultisetConstructionHoTT2019}. The proof of the
universal property of $\FSM$ is asserted by appealing to Mac Lane's coherence theorem for symmetric monoidal categories,
and using the fact that the finite symmetric group $\Sn$ encodes the group of permutations on a finite set. The existence of
the proof is folklore. We have produced a proof and formalised it in constructive type theory. A stronger universal
property that $\BFin$ is biinitial in $\term{RigCat}$, is in~\cite{elguetaGroupoidFiniteSets2021}.

### 2021-10-30 Pi/paper/finite.tex
Say symmetric group instead of permutation group

\subsection{Presenting the permutation group}

One way of thinking about presentations of $\Sn$ is via sorting algorithms, which use different primitive operations. A
sorting algorithm has to calculate a permutation of a list or a finite set, which satisfies the invariant of being a
sorted sequence, which means, the primitive operations of a sorting algorithm are able to generate all the permutations
on a given list. So, a chosen set of reversible operations in a sorting algorithm can be a good candidate for the
generators of a permutation group. For example, we could generate the permutation group on $\Fin[n]$ by using generators
(primitive operations) that:

@@

Finally, it is time to complete our goal of characterising the permutation groups. Having produced a Lehmer code by
normalising words in $\Sn$, we need to run it to produce a concrete bijection of finite sets, and, given a bijection
between finite sets, we need to encode it as a Lehmer code. We will prove that these maps construct an equivalence
between the types $\Lehmer[n]$ and $\Aut[\Fin[\suc[n]]]$~\footnote{Note that the indices for the type of permutations
  are off-by-one, because we chose $\Fin[n]$ to represent generators for permutations on $\Fin[\suc[n]]$.}. The idea for
this proof is borrowed from~\citet{Molzer-cubical}.

>>>>

\subsection{Presenting the symmetric group}

One way of thinking about presentations of $\Sn$ is via sorting algorithms, which use different primitive operations. A
sorting algorithm has to calculate a permutation of a list or a finite set, which satisfies the invariant of being a
sorted sequence, which means, the primitive operations of a sorting algorithm are able to generate all the permutations
on a given list. So, a chosen set of reversible operations in a sorting algorithm can be a good candidate for the
generators of a permutation group. For example, we could generate the symmetric group on $\Fin[n]$ by using generators
(primitive operations) that:

@@

Finally, it is time to complete our goal of characterising the symmetric group. Having produced a Lehmer code by
normalising words in $\Sn$, we need to run it to produce a concrete bijection of finite sets, and, given a bijection
between finite sets, we need to encode it as a Lehmer code. We will prove that these maps construct an equivalence
between the types $\Lehmer[n]$ and $\Aut[\Fin[\suc[n]]]$~\footnote{Note that the indices for the type of permutations
  are off-by-one, because we chose $\Fin[n]$ to represent generators for permutations on $\Fin[\suc[n]]$.}. The idea for
this proof is borrowed from~\citet{Molzer-cubical}.
